---
title: "2026 FIFA World Cup: Country Statistics and Visualizations"
author: "SICSS"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 120
)

library(tidyverse)
library(rvest)
library(janitor)
library(knitr)
library(scales)
library(countrycode)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

## Goal

This report pulls publicly available data on the **2026 FIFA World Cup participating countries** from Wikipedia, processes it with tidyverse, and produces:

1. A clean country-level summary table.
2. A bar chart of FIFA rankings among qualified teams.
3. A world map colored by FIFA ranking for those teams.

## 1) Pull data from Wikipedia

```{r pull-data}
# Wikipedia page for 2026 FIFA World Cup
wc_url <- "https://en.wikipedia.org/wiki/2026_FIFA_World_Cup"
wc_page <- read_html(wc_url)

# Read all tables and find the one containing qualified teams
all_tables <- wc_page %>% html_table(fill = TRUE)

qualified_raw <- all_tables %>%
  keep(~ any(str_detect(names(.x), regex("Team|Qualified", ignore_case = TRUE))) &&
         any(str_detect(unlist(.x[1:min(nrow(.x), 5), ], use.names = FALSE),
                        regex("Qualified teams|Hosts|AFC|CAF|UEFA|CONMEBOL|CONCACAF|OFC", ignore_case = TRUE)))) %>%
  .[[1]]

# Standardize names
qualified <- qualified_raw %>%
  clean_names()

qualified
```

## 2) Tidy and enrich data

```{r tidy-data}
# Attempt to identify columns robustly
team_col <- names(qualified)[str_detect(names(qualified), "team|nation|country")][1]
confed_col <- names(qualified)[str_detect(names(qualified), "confed|confederation|association")][1]
appearance_col <- names(qualified)[str_detect(names(qualified), "appearance")][1]
rank_col <- names(qualified)[str_detect(names(qualified), "rank")][1]

# If rank column is unavailable, scrape FIFA ranking from another Wikipedia page (simple fallback)
if (is.na(rank_col) || is.null(rank_col)) {
  fifa_rank_url <- "https://en.wikipedia.org/wiki/FIFA_Men%27s_World_Ranking"
  rank_page <- read_html(fifa_rank_url)
  rank_tables <- rank_page %>% html_table(fill = TRUE)

  rank_table <- rank_tables %>%
    keep(~ any(str_detect(names(.x), regex("Team|Country|Association", ignore_case = TRUE))) &&
           any(str_detect(names(.x), regex("Rank", ignore_case = TRUE)))) %>%
    .[[1]] %>%
    clean_names()

  rank_team_col <- names(rank_table)[str_detect(names(rank_table), "team|country|association")][1]
  rank_rank_col <- names(rank_table)[str_detect(names(rank_table), "rank")][1]

  rank_df <- rank_table %>%
    transmute(
      team_name = .data[[rank_team_col]] %>% as.character() %>% str_remove_all("\\[.*?\\]") %>% str_trim(),
      fifa_rank = parse_number(.data[[rank_rank_col]])
    ) %>%
    filter(!is.na(fifa_rank), !is.na(team_name), team_name != "")

  wc_clean <- qualified %>%
    transmute(
      team_name = .data[[team_col]] %>% as.character() %>% str_remove_all("\\[.*?\\]") %>% str_trim(),
      confederation = if (!is.na(confed_col)) .data[[confed_col]] else NA_character_,
      appearances = if (!is.na(appearance_col)) parse_number(.data[[appearance_col]]) else NA_real_
    ) %>%
    filter(!is.na(team_name), team_name != "") %>%
    distinct() %>%
    left_join(rank_df, by = "team_name")

} else {
  wc_clean <- qualified %>%
    transmute(
      team_name = .data[[team_col]] %>% as.character() %>% str_remove_all("\\[.*?\\]") %>% str_trim(),
      confederation = if (!is.na(confed_col)) .data[[confed_col]] else NA_character_,
      appearances = if (!is.na(appearance_col)) parse_number(.data[[appearance_col]]) else NA_real_,
      fifa_rank = parse_number(.data[[rank_col]])
    ) %>%
    filter(!is.na(team_name), team_name != "") %>%
    distinct()
}

# Harmonize common country names for mapping and joins
wc_clean <- wc_clean %>%
  mutate(
    team_name = recode(team_name,
      "United States" = "United States",
      "USA" = "United States",
      "Korea Republic" = "South Korea",
      "IR Iran" = "Iran",
      "Côte d'Ivoire" = "Ivory Coast"
    ),
    iso3c = countrycode(team_name, origin = "country.name", destination = "iso3c")
  )

wc_clean
```

## 3) Country summary table

```{r table}
summary_table <- wc_clean %>%
  arrange(fifa_rank) %>%
  select(team_name, confederation, appearances, fifa_rank)

kable(
  summary_table,
  caption = "Qualified 2026 World Cup teams: confederation, appearances, and FIFA ranking",
  align = c("l", "l", "r", "r")
)
```

## 4) Visualization 1: Bar chart of FIFA ranks

```{r viz-bar}
viz_data <- wc_clean %>%
  filter(!is.na(fifa_rank)) %>%
  arrange(fifa_rank) %>%
  mutate(team_name = fct_reorder(team_name, -fifa_rank))

ggplot(viz_data, aes(x = team_name, y = fifa_rank, fill = confederation)) +
  geom_col() +
  coord_flip() +
  scale_y_reverse() +
  labs(
    title = "FIFA rankings of 2026 World Cup participating teams",
    x = "Country",
    y = "FIFA Rank (lower is better)",
    fill = "Confederation"
  ) +
  theme_minimal()
```

## 5) Visualization 2: World map colored by FIFA rank

```{r viz-map}
world <- ne_countries(scale = "medium", returnclass = "sf")

map_data <- world %>%
  left_join(wc_clean %>% select(iso3c, team_name, fifa_rank), by = c("iso_a3" = "iso3c"))

ggplot(map_data) +
  geom_sf(aes(fill = fifa_rank), color = "grey70", size = 0.1) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    na.value = "grey92",
    labels = label_number()
  ) +
  labs(
    title = "2026 World Cup countries on world map",
    subtitle = "Colored by FIFA ranking (NA = not in participating list)",
    fill = "FIFA Rank"
  ) +
  theme_minimal()
```

## Notes

- Source page for teams: <https://en.wikipedia.org/wiki/2026_FIFA_World_Cup>
- Rankings are taken from the same table when available, with a fallback to FIFA ranking page scraping.
- This document is designed to render as GitHub-flavored markdown.
